---
title: "Setting up Clinical Studies with gsm.datasim"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Setting up Clinical Studies with gsm.datasim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup, eval = TRUE}
library(gsm.datasim)
```

# Introduction

The `gsm.datasim` package provides a comprehensive framework for generating synthetic clinical trial data. This vignette demonstrates how to set up both single snapshot and longitudinal studies, covering the key functions and configuration options available.

## Overview of Study Types

The package supports two main types of studies:

1. **Single Snapshot Studies**: Generate data for a single time point
2. **Longitudinal Studies**: Generate data across multiple time points to simulate study progression

# Single Snapshot Studies

A single snapshot study generates clinical data for one specific time point. This is useful for testing analytics workflows or when you need a static dataset.

## Basic Single Snapshot Setup

The simplest way to create a single snapshot study is using the study configuration system:

```{r single-basic}
# Create a basic study configuration
config <- create_study_config(
  study_id = "STUDY001",
  participant_count = 100,
  site_count = 5
)

# Add clinical domains (datasets)
config <- add_dataset_config(config, "Raw_AE", enabled = TRUE)     # Adverse Events
config <- add_dataset_config(config, "Raw_LB", enabled = TRUE)     # Laboratory Data
config <- add_dataset_config(config, "Raw_VISIT", enabled = TRUE)  # Visit Data

# Generate the study data
raw_data <- generate_study_data(config, verbose = TRUE)

# Examine the generated data structure
names(raw_data)
```

## Configuring Temporal Settings

Even for single snapshots, you can control the temporal aspects:

```{r single-temporal}
# Create configuration with specific temporal settings
config <- create_study_config(
  study_id = "STUDY002", 
  participant_count = 200,
  site_count = 10
)

# Set temporal configuration
config <- set_temporal_config(
  config,
  start_date = "2023-06-01",
  snapshot_count = 1,           # Single snapshot
  snapshot_width = "months"
)

# Add domains with custom configurations
config <- add_dataset_config(config, "Raw_AE", enabled = TRUE)
config <- add_dataset_config(config, "Raw_LB", enabled = TRUE, 
                            growth_pattern = "exponential")

# Generate data
raw_data <- generate_study_data(config, verbose = TRUE)
```

## Adding Analytics Pipeline

You can run analytics on your single snapshot study:

```{r single-analytics}
# Create study with analytics
config <- create_study_config(
  study_id = "STUDY003",
  participant_count = 150,
  site_count = 8,
  analytics_package = "gsm.mapping"  # or other analytics package
)

# Add clinical domains
config <- add_dataset_config(config, "Raw_AE", enabled = TRUE)
config <- add_dataset_config(config, "Raw_LB", enabled = TRUE)

# Generate data
raw_data <- generate_study_data(config, verbose = TRUE)

# Run analytics pipeline
analytics_results <- generate_analytics_layers(raw_data, config, verbose = TRUE)

# Examine analytics results
str(analytics_results, max.level = 2)
```

# Longitudinal Studies  

Longitudinal studies simulate data evolution over time, which is essential for understanding how clinical metrics change throughout a study.

## Basic Longitudinal Study

The `create_longitudinal_study()` function provides the easiest way to set up a longitudinal study:

```{r longitudinal-basic}
# Create a longitudinal study with default settings
study <- create_longitudinal_study(
  study_id = "LONG-001",
  participants = 100,
  sites = 5,
  snapshots = 6,                    # 6 time points
  interval = "1 month",             # Monthly snapshots
  domains = c("AE", "LB", "VISIT"), # Clinical domains to include
  include_pipeline = TRUE,          # Run analytics
  verbose = TRUE
)

# Examine the study structure
names(study)

# View snapshot dates
names(study$raw_data)

# Check analytics results (if pipeline was run)
if (!is.null(study$analytics)) {
  names(study$analytics)
}
```

## Advanced Longitudinal Configuration

For more control over the longitudinal study setup:

```{r longitudinal-advanced}
# Create a complex longitudinal study
study <- create_longitudinal_study(
  study_id = "ADVANCED-001",
  participants = 500,
  sites = 25,
  snapshots = 12,                               # 1 year of monthly data
  interval = "1 month",
  domains = c("AE", "LB", "VISIT", "PD", "PK"), # Multiple domains
  include_pipeline = TRUE,
  analytics_package = "gsm.kri",           # Specify analytics package
  verbose = TRUE
)

# Generate study summary
summary_data <- summarize_longitudinal_study(study)
print(summary_data)
```

## Custom Base Dates

You can specify a custom starting date for your longitudinal study:

```{r longitudinal-custom-date}
# Use the lower-level function for more control
mappings <- ensure_core_mappings(c("AE", "LB", "VISIT"))

raw_data <- generate_study_snapshots(
  study_id = "CUSTOM-001",
  participants = 100,
  sites = 8,
  snapshots = 6,
  interval = "2 months",                        # Bi-monthly snapshots
  mappings = mappings,
  base_date = "2022-01-15",                    # Custom start date
  verbose = TRUE
)

# Create study object
config <- list(
  participants = 100,
  sites = 8,
  snapshots = 6,
  interval = "2 months",
  domains = c("AE", "LB", "VISIT")
)

study <- create_longitudinal_study_data(
  study_id = "CUSTOM-001",
  raw_data = raw_data,
  config = config
)
```

## Different Time Intervals

The package supports various time intervals for longitudinal studies:

```{r longitudinal-intervals}
# Weekly snapshots
weekly_study <- create_longitudinal_study(
  study_id = "WEEKLY-001",
  participants = 50,
  sites = 3,
  snapshots = 12,
  interval = "1 week",              # Weekly intervals
  domains = c("AE", "VISIT"),
  verbose = TRUE
)

# Quarterly snapshots  
quarterly_study <- create_longitudinal_study(
  study_id = "QUARTERLY-001",
  participants = 300,
  sites = 15,
  snapshots = 8,
  interval = "3 months",            # Quarterly intervals  
  domains = c("AE", "LB", "PD"),
  verbose = TRUE
)
```

## Quick Study Creation

For rapid prototyping, use the `quick_longitudinal_study()` function:

```{r quick-study}
# Create a study with sensible defaults
quick_study <- quick_longitudinal_study(
  study_name = "QUICK-PROTO-001",
  participants = 200,
  sites = 10, 
  months_duration = 18,             # 18-month study
  study_type = "standard",
  run_analytics = TRUE,
  verbose = TRUE
)

# The function automatically determines appropriate snapshots and intervals
str(quick_study$config)
```

# Study Configuration Best Practices

## Domain Selection

Choose clinical domains based on your study objectives:

```{r domain-selection}
# Oncology study domains
oncology_domains <- c("AE", "LB", "PD", "OverallResponse", 
                     "VISIT", "DATACHG", "QUERY")

# Safety study domains  
safety_domains <- c("AE", "LB", "VISIT", "EXCLUSION",
                   "Death", "STUDCOMP")

# PK/PD study domains
pkpd_domains <- c("PK", "PD", "LB", "VISIT", "AE")

# Create study with oncology focus
onc_study <- create_longitudinal_study(
  study_id = "ONCOLOGY-001",
  participants = 300,
  sites = 20,
  snapshots = 24,
  interval = "1 month", 
  domains = oncology_domains,
  verbose = TRUE
)
```

## Analytics Configuration

Configure analytics based on your monitoring needs:

```{r analytics-config}
# Create study with specific analytics workflows
study <- create_longitudinal_study(
  study_id = "MONITORED-001",
  participants = 400,
  sites = 30,
  snapshots = 12,
  interval = "1 month",
  domains = c("AE", "LB", "VISIT", "QUERY"),
  include_pipeline = TRUE,
  analytics_package = "gsm.kri",
  analytics_workflows = c("kri0001", "kri0002"), # Specific workflows
  verbose = TRUE
)

# Examine specific analytics results
if (!is.null(study$analytics)) {
  # Site-level results
  site_results <- study$analytics$by_site$results
  
  # Study-level results  
  study_results <- study$analytics$by_study$results
}
```

# Data Examination and Validation

## Exploring Generated Data

```{r data-exploration}
# After generating a study, examine the data structure
study <- create_longitudinal_study(
  study_id = "EXPLORE-001",
  participants = 50,
  sites = 3,
  snapshots = 3,
  interval = "1 month",
  domains = c("AE", "LB"),
  verbose = TRUE,
  include_pipeline = FALSE
)

# Check data for each snapshot
for (i in 1:length(study$raw_data)) {
  snapshot_name <- names(study$raw_data)[i]
  snapshot_data <- study$raw_data[[i]]
  
  cat("Snapshot:", snapshot_name, "\n")
  cat("Datasets:", paste(names(snapshot_data), collapse = ", "), "\n")
  
  # Check sample sizes
  if ("Raw_SUBJ" %in% names(snapshot_data)) {
    cat("Subjects:", nrow(snapshot_data$Raw_SUBJ), "\n")
  }
  if ("Raw_AE" %in% names(snapshot_data)) {
    cat("AE Records:", nrow(snapshot_data$Raw_AE), "\n") 
  }
  cat("\n")
}
```

## Data Quality Checks

```{r data-quality}
# Function to validate study data
validate_study_data <- function(study) {
  validation_results <- list()
  
  for (snapshot_name in names(study$raw_data)) {
    snapshot <- study$raw_data[[snapshot_name]]
    
    # Check required datasets
    required_datasets <- c("Raw_STUDY", "Raw_SITE", "Raw_SUBJ", "Raw_ENROLL")
    missing_required <- setdiff(required_datasets, names(snapshot))
    
    validation_results[[snapshot_name]] <- list(
      datasets_present = names(snapshot),
      missing_required = missing_required,
      total_subjects = if ("Raw_SUBJ" %in% names(snapshot)) nrow(snapshot$Raw_SUBJ) else 0,
      total_sites = if ("Raw_SITE" %in% names(snapshot)) nrow(snapshot$Raw_SITE) else 0
    )
  }
  
  return(validation_results)
}

# Validate a study
validation <- validate_study_data(study)
str(validation)
```

# Conclusion

The `gsm.datasim` package provides flexible tools for creating both single snapshot and longitudinal clinical studies. Key takeaways:

- Use `create_study_config()` and `generate_study_data()` for single snapshots
- Use `create_longitudinal_study()` for multi-snapshot studies  
- Configure domains based on your study type and monitoring needs
- Leverage the analytics pipeline for automated quality monitoring
- Use `quick_longitudinal_study()` for rapid prototyping

The modular design allows you to start simple and add complexity as needed, making it suitable for both exploratory work and production data generation scenarios.